\pkg{cleanR}: Maid for Cleaning Datasets in \proglang{R}

### 2016-12-20 latex/article_vol2.tex
Text wrangling

  We describe an \proglang{R} package which implements an extensive
  and customizeable suite of quality assessment tools to be applied to
  a dataset in order to identify potential problems in its
  variables. The results can be presented in an auto-generated,
  non-technical, stand-alone overview document, intended to be perused
  by an investigator with an understanding of the variables in the
  data and the experimental design, but not necessarily knowledge of
  \R{R}. Thereby, \pkg{cleanR} aids the dialogue between data analysist
  and field experts, while also providing easy documentation of data
  cleaning steps and data quality control. Moreover, the \pkg{cleanR}
  solution changes the data cleaning process from the usual ad hoc
  approach to a systematic, well-documented endeavor.  \pkg{cleanR} also
  provides a suite of more typical \proglang{R} tools for interactive data
  quality assessment and -cleaning. \hl{Hvad henvises der
    til  med denne sætning?}

@@

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and on data wrangling. Online tools such as OpenRefine
(\url{http://openrefine.org/}) and \proglang{R}-packages such as
\pkg{data.table}, and \pkg{plyr} have made data wrangling a lot
easier, but only a handful of packages such as \pkg{editrules},
\pkg{validate}, \pkg{DataCombine}, and \pkg{janitor} attempt to
implement systematic, reproducible data cleaning.  These packages use
different approaches for data cleaning: \pkg{editrules} and
\pkg{validate} provide frameworks for setting up and checking
constraints on the variables, while \pkg{DataCombine} and
\pkg{janitor} both provide a few functions for identifying problems
(e.g, duplicates, dates coded as numbers) in data.

While some of these tools ...

data cleaning is usually done in a very ad hoc manner

 checking your data easy, maintainable and reproducibl

\hl{What are these tools? More here about the solutions
  already available.}

\hl{A text here (with offset in what tools are already available) that
\begin{itemize}
\item Makes the point that data cleaning is usually done in a very ad hoc manner
\item Maybe hints to the discussion later about poor documentation of data cleaning/quality assessment as the status quo for most analysts
\item Gives a specification of what we mean by data errors/mistakes/whatever. I'm not completely sure this will be clear to everyone and we should probably mention it early on to not lose readers/users.
\end{itemize}
}

Data cleaning is a time consuming endeavor, as it inherently requires
human interaction since every dataset is different and the variables
in the dataset can only be understood in the proper context of their
origin.  \hl{Segway: Something like maybe this is why proper data
  cleaning is not always done...}  In many situations these errors are
discovered in the process of the data analysis (e.g., \hl{a
  categorical variable with numeric labels for each category may be 
  wrongly classified as a quantitative variable\footnote{\hl{Is that
      something we can fix in clean()? Not quite sure what "labels"
      mean in this context. --- Kun i den forstand, at den skriver
      klassen ud, så ja. Har prøvet at opdatere teksten}}} or a
variable where all values have erroneously been coded to the same
value), but in other cases a human with knowledge about the data
context area is needed to identify possible mistakes in the data
(e.g., if there are 4 categories for a
variable that should only have 3).  \hl{segway} \\

The \pkg{cleanR} approach to data cleaning and -quality assessment is characterized by two principles \hl{use a different term than "principles" here and below, when we refer to this sentence}. First of all, there is no need for data cleaning to be an ad hoc procedure. Often, we have a very clear idea of what flags are raisable in a given dataset before we look at it, as we were the ones to produce it in the first place. This means that data cleaning can easily be a well-documented, well-specified procedure. In order to aid this principle, \pkg{cleanR} provides easy-to-use, automated tools for data quality assessment in \R{R} on which data cleaning decisions can be build. This quality assessment is presented in a auto-generated overview document, readable by data analysts and field experts alike, thereby also contributing to a inter-field dialogue about the data at hand. Oftentimes, e.g. distinguishing between faulty codings of a numeric value and unusual, but correct, values requires problem-specific expertise that might not be held by the data analyst. Hopefully, having easy access to data descriptions through \pkg{cleanR} will help this necessary knowledge sharing.

While \pkg{cleanR}s primary raison d'être is auto-generating data quality assessment overview documents, we still wish to emphasize that it is \textit{not} a tool for automatic data cleaning. This qualifies as the second principle of \pkg{cleanR}: Data cleaning decisions should always be made by humans. Therefore, \pkg{cleanR} does not supply any tools for "fixing" errors in the data. However, we do provide interactive functions that can be used to identify potentially erroneous entries in a dataset and that can make it easier to solve data issues, one variable at a time.\\

This manuscript is structured as follows: First, in section \hl{2}, we introduce the representative of the first \hl{principle}, namely the \R{clean()} function, which generates data cleaning overview documents. In the \pkg{cleanR} package, we have provided a number of default cleaning steps that cover the data cleaning challenges, we find to be most common. Next, in section \hl{3}, we present the interactive mode of \pkg{cleanR}, as motivated by the second \hl{principle} above. But, as any data analyst knows, every dataset is different, and some datasets might include problems that cannot be detected by our data checking functions. In Section \hl{4}, we turn to the question of how such \pkg{cleanR} extensions can be made, such that they are integrable with the \R{clean()} function and with the other tools available in \pkg{cleanR}.  At last, in Section 5 \hl{proper ref}, we discuss a number of examples of specific data cleaning challenges and how \pkg{cleanR} can be used to solve them. \hl{Is there a Section 6?}\\

\hl{Do they typically discuss notation in articles like this? I try (somewhat inconsistently) to refer to functions as \R{function()} while other \R{R} objects are referred to merely as \R{object}. This is also the style adopted by Wickham in his books.}

@@

\section{Something like "producing a data cleaning overview"} \label{sec:example1}

\begin{figure}
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{cleanR_toyData.pdf}} 
\frame{\includegraphics[width=7.5cm, page=3]{cleanR_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{cleanR_testData.pdf}
\end{center}
\label{fig:example1}
\caption{Example output from running \R{clean()} on the \R{toyData} dataset. First a summary of the full dataset,.. 
  \hl{Maybe fix these pdfs such that they are one file with no blankspace? I.e. basically the .html-version of this file printed as pdf. This version is too small, right?}}
\end{figure}

@@

The \R{clean()} function is the primary workhorse of \pkg{cleanR} and
this is the only function needed by the user if the standard battery
of tests are used to output the data cleaning summaries. The data
cleaning output itself is an overview document, intended for reading
by humans, in either pdf or html format. Appendix \hl{something}
provides an example of a data cleaning output document, produced by
calling \R{clean()} on the dataset \R{toyData} available in
\pkg{cleanR}. The first two pages of this data cleaning output are
also available in Figure~\ref{fig:example1}. \R{toyData} is a very
small ($15$ rows and $6$ variables), artificial dataset, whose only purpose is to
illustrate the main capabilities of \pkg{cleanR}. The following
commands load the dataset and produce the cleaning output:

By default, a pdf overview document is produced, saved to the disc (in
the working directory) and opened for immediate inspection. Turning to
Figure~\ref{fig:example1}, we see that such a data cleaning output
document consists of two parts. First, an overview of what was done is
presented under the title \textit{Data cleaning summary}. Secondly,
each variable in the dataset is presented in turn using (up to) three
tools in the \textit{Variable list}: A table summarizing key features
of the variable, a figure visualizing its distribution and potentially
also a list of flagged issues. For instance, in the \R{numeric}-type
variable \R{var2} from \R{toyData}, \R{clean()} has identified two
values that are suspected to be miscoded missing values (\R{999} and
\R{NaN}), while two values were also flagged as potential outliers
that should be investigated more carefully. We can then return to Part
1, the data cleaning summary, and inspect what sorts of checks were
performed on this variable. \hl{hm, delete the last sentence? or use
  different variable? Only these exact two checks were performed}.

Though the \R{clean()} function is very easy to use, it should not be mistaken to be inflexible. A large number of function arguments allows for the cleaning overview document to be molded as the user sees best. Below, we present a selection of these arguments in sections \hl{xx} and \hl{xxx}. We do this under two distinct headlines to emphasize that two levels of control are available when using \R{clean()}. Either, we can control the \hl{[something] - essentially, this is everything but SVC-functions - } or we can control what quality assessments and descriptions each variable are subjected to. In order to understand this distinction, a brief glimpse of the inner structure of \R{clean()} might be useful. This can be obtained from Figure~\ref{figure:cleanStructure}. \hl{segway to next bit or just something to end the paragraph}

@@

\section{Appendix Something}
\includepdf[pages=2-4, pagecommand={}, frame=true, noautoscale=true, scale=0.7]{cleanR_toyData.pdf}

>>>>

  We describe an \proglang{R} package which implements an extensive
  and customizeable suite of quality assessment tools to be applied to
  a dataset in order to identify potential problems in its
  variables. The results can be presented in an auto-generated,
  non-technical, stand-alone overview document, intended to be perused
  by an investigator with an understanding of the variables in the
  data and the experimental design, but not necessarily knowledge of
  \R{R}. Thereby, \pkg{cleanR} aids the dialogue between data
  analysist and field experts, while also providing easy
  documentation of reproducible data cleaning steps and data quality
  control. Moreover, the \pkg{cleanR} solution changes the data
  cleaning process from the usual ad hoc approach to a systematic,
  well-documented endeavor.  \pkg{cleanR} also provides a suite of
  more typical \proglang{R} tools for interactive data quality
  assessment and -cleaning. \hl{Hvad henvises der til med denne
    sætning?}

@@

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and on data wrangling. While a computer should never
provide an unsupervised decision on what should be done to potential
errors in the dataset, it can be an extremely useful tool in tracking
down and flagging potential erroneous data and in providing
information for humans to easier identify errors based on the context.

Online tools such as OpenRefine (\url{http://openrefine.org/}) and
\proglang{R}-packages such as \pkg{data.table}, and \pkg{plyr} have
made data wrangling a lot easier, but only a handful of packages such
as \pkg{editrules}, \pkg{validate}, \pkg{DataCombine}, and
\pkg{janitor} attempt to implement systematic, reproducible data
cleaning.  These packages use different approaches for data cleaning:
\pkg{editrules} and \pkg{validate} provide frameworks for setting up
and checking constraints on the variables, while \pkg{DataCombine} and
\pkg{janitor} both provide a few functions for identifying problems
(e.g, duplicates, dates coded as numbers, etc.) in data.

While these tools attempt to alleviate the ubiquitous ad hoc approach
to data cleaning they are primarily intended for the data savvy users
and less on the general researcher with a knowledge about the specific
field and context of the available data. The \pkg{cleanR} package
tries to address this by providing a framework that both allows for
extendable, systematic, reproducible data cleaning, and summarizing
findings for researchers from other fields such that they can act as
human experts when tracking down potential errors.

% \begin{itemize}
% \item Makes the point that data cleaning is usually done in a very ad hoc manner
% \item Maybe hints to the discussion later about poor documentation of data cleaning/quality assessment as the status quo for most analysts
% \item Gives a specification of what we mean by data errors/mistakes/whatever. I'm not completely sure this will be clear to everyone and we should probably mention it early on to not lose readers/users.
% \end{itemize}
% }

Data cleaning is a time consuming endeavor, as it inherently requires
human interaction since every dataset is different and the variables
in the dataset can only be understood in the proper context of their
origin. This often requires a collaborative effort between an expert
in the field and a statistician or data scientist, which may be why
the process of proper data cleaning is not always undertaken. In many
situations, these errors are discovered in the process of the data
analysis (e.g., a categorical variable with numeric labels for each
category may be wrongly classified as a quantitative variable or a
variable where all values have erroneously been coded to the same
value), but in other cases a human with knowledge about the data
context area is needed to identify possible mistakes in the data
(e.g., if there are 4 categories for a variable that should only have
3).

The \pkg{cleanR} approach to data cleaning and -quality assessment is
governed by two fundamental paradigms. First of all, there is no need
for data cleaning to be an ad hoc procedure. Often, we have a very
clear idea of what flags are raisable in a given dataset before we
look at it, as we were the ones to produce it in the first place. This
means that data cleaning can easily be a well-documented,
well-specified procedure. In order to aid this paradigm, \pkg{cleanR}
provides easy-to-use, automated tools for data quality assessment in
\proglang{R} on which data cleaning decisions can be build. This quality
assessment is presented in an auto-generated overview document,
readable by data analysts and field experts alike, thereby also
contributing to a inter-field dialogue about the data at
hand. Oftentimes, e.g. distinguishing between faulty codings of a
numeric value and unusual, but correct, values requires
problem-specific expertise that might not be held by the data
analyst. Hopefully, having easy access to data descriptions through
\pkg{cleanR} will help this necessary knowledge sharing.

While \pkg{cleanR}s primary raison d'être is auto-generating data
quality assessment overview documents, we still wish to emphasize that
it is \emph{not} a tool for unsupervised data cleaning. This qualifies
as the second paradigm of \pkg{cleanR}: Data cleaning decisions
should always be made by humans. Therefore, \pkg{cleanR} does not
supply any tools for ``fixing'' errors in the data. However, we do
provide interactive functions that can be used to identify potentially
erroneous entries in a dataset and that can make it easier to solve
data issues, one variable at a time.

This manuscript is structured as follows: First, in Section
\ref{sec:usingcleanR}, we introduce the representative of the first
paradigm, namely the \R{clean()} function, which generates data
cleaning overview documents. In the \pkg{cleanR} package, we have
provided a number of default cleaning steps that cover the data
cleaning challenges, we find to be most common. Next, in Section
\hl{3}, we present the interactive mode of \pkg{cleanR}, as motivated
by the second paradigm above. But, as any data analyst knows,
every dataset is different, and some datasets might include problems
that cannot be detected by our data checking functions. In Section
\hl{4}, we turn to the question of how such \pkg{cleanR} extensions
can be made, such that they are integrable with the \R{clean()}
function and with the other tools available in \pkg{cleanR}.  At last,
in Section 5 \hl{proper ref}, we discuss a number of examples of
specific data cleaning challenges and how \pkg{cleanR} can be used to
solve them.

% \hl{Is there a Section 6?}

%\hl{Do they typically discuss notation in articles like this? I try (somewhat inconsistently) to refer to functions as \R{function()} while other \R{R} objects are referred to merely as \R{object}. This is also the style adopted by Wickham in his books.}

@@

\section{Creating a data cleaning overview} \label{sec:example1}\label{sec:usingcleanR}

The \R{clean()} function is the primary workhorse of \pkg{cleanR} and
this is the only function needed by the user if the standard battery
of tests are used to output the data cleaning summaries. The data
cleaning output itself is an overview document, intended for reading
by humans, in either pdf or html format. Appendix \ref{sec:appendix1}
provides an example of a data cleaning output document, produced by
calling \R{clean()} on the dataset \R{toyData} available in
\pkg{cleanR}. The first two pages of this data cleaning output are
shown in Figure~\ref{fig:example1}. \R{toyData} is a very
small ($15$ rows and $6$ variables), artificial dataset, whose only purpose is to
illustrate the main capabilities of \pkg{cleanR}. The following
commands load the dataset and produce the cleaning output:

\begin{figure}
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{cleanR_toyData.pdf}} 
\frame{\includegraphics[width=7.5cm,page=3]{cleanR_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{cleanR_testData.pdf}
\end{center}
\label{fig:example1}
\caption{Example output from running \R{clean()} on the \R{toyData}
  dataset. First a summary of the full dataset is given and then
  type-dependent information on each variable is given in a table and
  a graph. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\end{figure}

By default, a pdf overview document is produced, saved to the disc (in
the working directory) and opened for immediate inspection. Turning to
Figure~\ref{fig:example1}, we see that such a data cleaning output
document consists of two parts. First, an overview of what was done is
presented under the title \textit{Data cleaning summary}. Secondly,
each variable in the dataset is presented in turn using (up to) three
tools in the \textit{Variable list}: A table summarizing key features
of the variable, a figure visualizing its distribution and potentially
also a list of flagged issues. For instance, in the \R{numeric}-type
variable \R{var2} from \R{toyData}, \R{clean()} has identified two
values that are suspected to be miscoded missing values (\R{999} and
\R{NaN}), while two values were also flagged as potential outliers
that should be investigated more carefully. 
%We can then return to Part
%1, the data cleaning summary, and inspect what sorts of checks were
%performed on this variable. \hl{hm, delete the last sentence? or use
%  different variable? Only these exact two checks were performed}.

Though the \R{clean()} function is very easy to use, it should not be
mistaken to be inflexible. A large number of function arguments allows
for the cleaning overview document to be molded according to the
user's needs. The most common arguments are summarized in Table~\ref{}
but in order to understand this distinction, a brief glimpse of the
inner structure of \R{clean()} might be useful. This can be obtained
from Figure~\ref{figure:cleanStructure}.

Below, we present a selection of these arguments in
sections \hl{xx} and \hl{xxx}. We do this under two distinct headlines
to emphasize that two levels of control are available when using
\R{clean()}. Either, we can control the \hl{[something] - essentially,
  this is everything but SVC-functions - } or we can control what
quality assessments and descriptions each variable are subjected
to.  \hl{segway to next bit or
  just something to end the paragraph}

@@

@@

\section{Appendix Something} \label{sec:appendix1}
\includepdf[pages=2-4, pagecommand={}, frame=true, noautoscale=true, scale=0.7]{cleanR_toyData.pdf}
